---
title: "Assignment 4: Data Wrangling"
author: "Yoyo Shi"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  markdown: 
    wrap: 72
---

## OVERVIEW

In this assignment, you will build a reproducible workflow that cleans
four EPA air-quality files, combines them, reshapes the result, and
produces grouped summaries. Date conversion is a brief review from
Module 3; the emphasis is on data wrangling.

## Directions

1.  Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing
    `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill
    each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code
    into a single PDF file.
6.  Ensure that code in code chunks does not extend off the page in the
    PDF.

## Set up your session

### Questions 1–2

1a. Load the `tidyverse` and `here` packages into your session.

1b. Check your working directory.

1c. Read all four raw files associated with the EPA Air dataset. See the
README file for information about the variables and measurements.

2.  Add the appropriate code to reveal the dimensions of the four
    datasets.

```{r setup questions 1-2, message = FALSE}
#1a 
library(tidyverse)
library(here)

#1b 
here()

#1c 
EPAair_O3_NC2024_raw <- read.csv("Data/Raw/EPAair_O3_NC2024_raw.csv")
EPAair_O3_NC2025_raw <- read.csv("Data/Raw/EPAair_O3_NC2025_raw.csv")
EPAair_PM25_NC2024_raw <- read.csv("Data/Raw/EPAair_PM25_NC2024_raw.csv")
EPAair_PM25_NC2025_raw <- read.csv("Data/Raw/EPAair_PM25_NC2025_raw.csv")

#2 
dim(EPAair_O3_NC2024_raw)
dim(EPAair_O3_NC2025_raw)
dim(EPAair_PM25_NC2024_raw)
dim(EPAair_PM25_NC2025_raw)

```

> All four datasets should have the same number of columns (PM2.5 has
> one extra column), but unique record counts (rows). Do your datasets
> follow this pattern? Answer: Yes

## Wrangle individual datasets to create processed files.

### Questions 3–5

3.  Change the Date columns to be date objects.

4.  Select the following columns: DDate, Daily.AQI.Value,
    Local.Site.Name, AQS.Parameter.Description, County, Site.Latitude,
    Site.Longitude

5.  For the PM2.5 datasets, fill all cells in AQS.Parameter.Description
    with "PM2.5" (all cells in this column should be identical).

```{r questions 3 thru 5}
#3
EPAair_O3_NC2024_raw$Date <- mdy(EPAair_O3_NC2024_raw$Date)
EPAair_O3_NC2025_raw$Date <- mdy(EPAair_O3_NC2025_raw$Date)
EPAair_PM25_NC2024_raw$Date <- mdy(EPAair_PM25_NC2024_raw$Date)
EPAair_PM25_NC2025_raw$Date <- mdy(EPAair_PM25_NC2025_raw$Date)

#4
EPAair_O3_NC2024_processed <- select(EPAair_O3_NC2024_raw, Date, 
                                     Daily.AQI.Value, Local.Site.Name,
                                     AQS.Parameter.Description, County, 
                                     Site.Latitude, Site.Longitude)
EPAair_O3_NC2025_processed <- select(EPAair_O3_NC2025_raw, Date, 
                                     Daily.AQI.Value, Local.Site.Name,
                                     AQS.Parameter.Description, County, 
                                     Site.Latitude, Site.Longitude)
EPAair_PM25_NC2024_processed <- select(EPAair_PM25_NC2024_raw, Date, 
                                     Daily.AQI.Value, Local.Site.Name,
                                     AQS.Parameter.Description, County, 
                                     Site.Latitude, Site.Longitude)
EPAair_PM25_NC2025_processed <- select(EPAair_PM25_NC2025_raw, Date, 
                                     Daily.AQI.Value, Local.Site.Name,
                                     AQS.Parameter.Description, County, 
                                     Site.Latitude, Site.Longitude)


#5
EPAair_PM25_NC2024_processed <- mutate(EPAair_PM25_NC2024_processed, 
                                       AQS.Parameter.Description = "PM2.5")
EPAair_PM25_NC2025_processed <- mutate(EPAair_PM25_NC2025_processed, 
                                       AQS.Parameter.Description = "PM2.5")

```

### Question 6

6.  Save all four processed datasets in the Processed folder. Use the
    same file names as the raw files but replace "raw" with "processed".

```{r question 6}
#6 
write.csv(
  EPAair_O3_NC2024_processed, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_NC2024_processed.csv")
)
write.csv(
  EPAair_O3_NC2025_processed, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_NC2025_processed.csv")
)
write.csv(
  EPAair_PM25_NC2024_processed, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_PM25_NC2024_processed.csv")
)
write.csv(
  EPAair_PM25_NC2025_processed, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_PM25_NC2025_processed.csv")
)
```

## Combine datasets

### Question 7

7.  Stack the four datasets with `bind_rows()`. Check that their column
    names are identical first.

```{r question 7}
#7 
#check column names
colnames(EPAair_O3_NC2024_processed)
colnames(EPAair_O3_NC2025_processed)
colnames(EPAair_PM25_NC2024_processed)
colnames(EPAair_PM25_NC2024_processed)

#stack
EPAair_combined <- bind_rows(EPAair_O3_NC2024_processed,
                             EPAair_O3_NC2025_processed,
                             EPAair_PM25_NC2024_processed,
                             EPAair_PM25_NC2024_processed)
EPAair_combined$Local.Site.Name <- as.factor(EPAair_combined$Local.Site.Name) 
```

### Question 8

8.  Use one pipeline (`%>%`) to create a daily site-level dataset that
    meets the following conditions:

-   Include only sites that the four data frames have in common:

"Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",\
"Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West
Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri.
Center", "Bryson City", "Millbrook School"

(the function `intersect` can figure out common factor levels - but it
will include sites with missing site information, which you don't
want...)

-   Some sites have multiple measurements per day. Group by date, site
    name, AQS parameter, and county, then calculate the mean AQI value,
    latitude, and longitude. Handle missing values explicitly in each
    mean.

-   Add numeric `Month` and `Year` columns from `Date`. Check if results
    are integers.

-   Hint: the dimensions of this dataset should be 15,012 × 9.

```{r question 8}
#8 
EPAair_daily <- 
  EPAair_combined %>% 
  filter(Local.Site.Name %in% c("Linville Falls", "Durham Armory", 
                                "Leggett", "Hattie Avenue", "Clemmons Middle",
                                "Mendenhall School", "Frying Pan Mountain", 
                                "West Johnston Co.", "Garinger High School", 
                                "Castle Hayne", "Pitt Agri. Center", 
                                "Bryson City", "Millbrook School")) %>% 
  group_by(Date, Local.Site.Name, AQS.Parameter.Description, County) %>% 
  filter(!is.na(Daily.AQI.Value) & 
           !is.na(Site.Latitude) & 
           !is.na(Site.Longitude)) %>%
  summarize(meanAQI = mean(Daily.AQI.Value),
            meanlatitude = mean(Site.Latitude),
            meanlongitude = mean(Site.Longitude),
            .groups = "drop") %>% 
  mutate(Month = month(Date), 
         Year = year(Date))

```

### Questions 9–10

9.  Use `pivot_wider()` so that ozone and PM2.5 AQI values occupy
    separate columns. Each location-date combination should occupy one
    row.

10. Call up the dimensions of your new tidy dataset.

```{r questions 9-10}
#9
EPAair_daily_wide<- pivot_wider(EPAair_daily, 
                                names_from = AQS.Parameter.Description, 
                                values_from = meanAQI)

#10
dim(EPAair_daily_wide)

```

### Question 11

11. Save your processed dataset with the following file name:
    "EPAair_O3_PM25_NC2425_Processed.csv"

```{r question 11}
#11
write.csv(
  EPAair_daily_wide, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_PM25_NC2425_Processed.csv")
)

```

## Generate summary tables

### Questions 12–13

12. Create a monthly summary data frame grouped by site, month, and
    year. Calculate mean ozone and PM2.5 AQI for each group, handling
    missing values explicitly. Then use `drop_na()` to remove rows only
    when mean **ozone** is missing. Mean PM2.5 may remain missing.

13. Call up the dimensions of the summary dataset.

```{r questions 12 and 13}
#12
EPAair_month_summary <-
  EPAair_daily_wide %>% 
  group_by(Local.Site.Name, Month, Year) %>% 
  summarize(mean_ozone = mean(Ozone),
            mean_pm2.5 = mean(PM2.5),
            .groups = "drop") %>% 
  drop_na(mean_ozone)

EPAair_month_summary2 <-
  EPAair_daily_wide %>% 
  group_by(Local.Site.Name, Month, Year) %>% 
  summarize(mean_ozone = mean(Ozone),
            mean_pm2.5 = mean(PM2.5),
            .groups = "drop") %>% 
  na.omit(mean_ozone)

#13
dim(EPAair_month_summary)
dim(EPAair_month_summary2)

```

### Question 14

14. Why is `drop_na(mean_ozone)` more appropriate here than `na.omit()`?
    Test both approaches and compare the dimensions of the resulting
    summary data frames.

> Answer: `na.omit` creates a dataframe with a dimension of `r dim(EPAair_month_summary2)` which is a lot shorter than when we use `drop_na(mean_ozone)` with a dimension of `r dim(EPAair_month_summary)`. This is because `na.omit` also removed the na from `mean_pm2.5` and we don't want that.
